Research Experience

### 2010-04-28 doc/thesis/thesis.tex
small amount of work on previous studies

\usepackage{times}
%\usepackage{tikz}
%\usetikzlibrary{trees}
%\usetikzlibrary{shapes,snakes}
\usepackage{verbatim}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{algorithmic}
\usepackage{setspace}
\usepackage{graphicx}
\usepackage{cancel}

@@

Because of the significant possility of a completely independent CMRO2,
extremely complex models for metabolism exist (\cite{Zheng2005}), although 
most recent studies have
focuse on their ability to explain the prolonged BOLD post stimulus 
undershoot \cite{Zheng2005}, \cite{Buxton2004}. This is because \cite{Buxton2004}
and later \cite{Riera2004} showed that the main portion of the signal may be 
accurately estimated by a simple blood flow locked expression of the CMRO2.. 
Although \cite{Deneux2006} did not deal extensively with prolonged post
stimulus undershoot, the comparisons made in that publication showed minimal
improvement from separate expressions of CMRO2, in comparison to the much 
increased complexity. \cite{Deneux2006} did show that by simply adding viscoelastic
terms, first proposed in \cite{Buxton2004}, that a slowed return to baseline for
the BOLD signal is possible to model. However, viscoelastic effects primarily
control CBV, which, as mentioned already, many studies have claimed cannot be 
responsible for the BOLD post-stimulus undershoot. Another extensive model
that attempts to quantify the post-stimulus undershoot is the compliance model 
proposed by \cite{Behzadi2005}. Alhough through a somewhat different means than
the \cite{Zheng2005} and \cite{Buxton2004} papers, its possible the increased
model flexibility ultimately is the key reason for the improvements, as opposed
to increased plausibility. Because of these controversies, and because this is
the first time a particle filter has been used for this problem in this way, 
our aim to keep the model simple was best met by using the original balloon
model with the possible addition of the visco-elastic effects from \cite{Buxton2004}.

Even more advanced versions of the Balloon model exist. In fact \cite{Buxton2004}
introduced several additional state variables, including the CMRO2, the O2 extraction
fraction, which is closely related to CMR02, and the neural response, which 
essentially starts high and decays toward some steady state value. The 
neural response is intended to emulate neural habituation, wherein neurons
become less sensitive to a prolonged stimulus. This goes back to the issue
of drastically increasing the model variance, for a minor decrease in model
bias, \cite{Deneux2006}. The parameters of the model have decently reliable
expected values by now. Although \cite{Buxton1998} and \cite{Friston2000}
both propose reasonable values for the model parameters, \cite{Friston2002} 
was the first paper to actually use an algorithm and test those parameters:
\begin{eqnarray}
epsilon &=& .54 \pm .1    \nonumber \\
tau_s & =& 1.54 \pm .25   \nonumber \\
tau_f & =& 2.46 \pm .25   \nonumber \\
tau_0 & =& .98 \pm .25    \nonumber \\
alpha & =& .33 \pm .45    \nonumber \\
E_0   & =& .34 \pm .1     \nonumber \\
V_0  & = & .03 (not estimated) \nonumber
\end{eqnarray}
later \cite{Hu2009} used an Unscented Kalman Filter to confirm these findings.
On the other hand \cite{Johnston2008} used a hybrid particle filter Expectation
Maximization algorithm to calculate parameters simultaneously with state.
His findings differed greatly from Friston et. al.:
\begin{eqnarray}
epsilon &=& .069 \pm .014    \nonumber \\
tau_s & =& 4.98 \pm 1.07  \nonumber \\
tau_f & =& 8.31 \pm 1.51   \nonumber \\
tau_0 & =& 8.38 \pm 1.5    \nonumber \\
alpha & =& .189 \pm .004   \nonumber \\
E_0   & =& .635 \pm .072     \nonumber \\
V_0   & =& .0149 \pm .006     \nonumber 
\end{eqnarray}
On the one hand, Johnston et. al. is a pseudo-maximum likelihood estimator, and 
it depends very little on the priors. On the other hand, the results are very 
different from \cite{Buxton1998}, \cite{Hu2009}, \cite{Friston2000}, and 
\cite{Friston2002}. Then again, the papers mentioned all have a common root,
and all used essentially the same priors, and all use Bayesian estimation
that depends greatly on the priors. It would be interesting to see what
effect different prior's would have on the results.

>>>>

\usepackage{times}
%\usepackage{tikz}
%\usetikzlibrary{trees}
%\usetikzlibrary{shapes,snakes}
\usepackage{verbatim}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{algorithmic}
\usepackage{setspace}
\usepackage{graphicx}
\usepackage{cancel}
\usepackage{hyperref}

@@

Because of the significant possility of a completely independent CMRO2,
extremely complex models for metabolism exist (\cite{Zheng2005}), although 
most recent studies have
focused on their ability to explain the prolonged BOLD post stimulus 
undershoot \cite{Zheng2005}, \cite{Buxton2004}. This is because \cite{Buxton2004}
and later \cite{Riera2004} showed that the main portion of the signal may be 
accurately estimated by a simple blood flow locked expression of the CMRO2.. 
Although \cite{Deneux2006} did not deal extensively with prolonged post
stimulus undershoot, the comparisons made in that publication showed minimal
improvement from separate expressions of CMRO2, in comparison to the much 
increased complexity. \cite{Deneux2006} did show that by simply adding viscoelastic
terms, first proposed in \cite{Buxton2004}, that a slowed return to baseline for
the BOLD signal is possible to model. However, viscoelastic effects primarily
control CBV, which, as mentioned already, many studies have claimed cannot be 
responsible for the BOLD post-stimulus undershoot. Another extensive model
that attempts to quantify the post-stimulus undershoot is the compliance model 
proposed by \cite{Behzadi2005}. Although through a somewhat different means than
the \cite{Zheng2005} and \cite{Buxton2004} papers, its possible the increased
model flexibility ultimately is the key reason for the improvements, as opposed
to increased plausibility. Because of these controversies, and because this is
the first time a particle filter has been used for this problem in this way, 
our aim to keep the model simple was best met by using the original balloon
model with the possible addition of the visco-elastic effects from \cite{Buxton2004}.

Even more advanced versions of the Balloon model exist. In fact \cite{Buxton2004}
introduced several additional state variables, including the CMRO2, the O2 extraction
fraction, which is closely related to CMR02, and the neural response, which 
causes the stimulus to decay toward some steady state value. The 
neural response is intended to emulate neural habituation, wherein neurons
become less sensitive to a prolonged stimulus. While these advanced may be more
capable of capturing a more exact version of the BOLD signal, the difference
between the model will often times be below the noise floor. In essence this 
is a classic bias-variance
dilemma: at some point increased model flexibility, and thus variance, is not
worth the decrease in model bias. For now it remains to be seen where this
line may be drawn in the BOLD signal, although \cite{deneux2006} does not
show a significant improvement from the additional parameters added in 
\cite{Buxton2004}.

\subsection{Previous Studies of Parameters}
There have been quite a few efforts to quantify the parameters of the
various BOLD models.  Although \cite{Buxton1998} and \cite{Friston2000}
both proposed physiologically reasonable values for the model parameters, 
\cite{Friston2002} was the first paper to calculate the parameters based 
on actual FMRI data. In that paper, Friston et. al. used a variation of
Expectation Maximization to find a normal distribution for the parameters:
\begin{eqnarray}
\epsilon &=& N(.54 , .1 ^2 )  \nonumber \\
\tau_s & =&  N(1.54, .25^2)   \nonumber \\
\tau_f & =&  N(2.46, .25^2)   \nonumber \\
\tau_0 & =&  N(.98 , .25^2 )   \nonumber \\
\alpha & =&  N(.33 , .45^2 )   \nonumber \\
E_0   & =&  N(.34 ,  .1 ^2 )   \nonumber \\
V_0  & = &  .03 (not\ estimated) \nonumber
\end{eqnarray}
Since then, several other methods of have been used to calculate
BOLD parameters from FMRI timecourses. In \cite{Riera2004}, a maximum
likelihood method for innovation processes was used as described by
\cite{Ozaki1994}
later \cite{Hu2009} used an Unscented Kalman Filter to confirm these findings.
Both \cite{Hu2009} and \cite{Friston2002} came to results very similar to the
expected values.
On the other hand \cite{Johnston2008} used a hybrid particle filter Expectation-
Maximization algorithm to calculate parameters and state simultaneously.
His derived parameter distributions differed greatly from Friston et. al.:
\begin{eqnarray}
epsilon &=& .069 \pm .014    \nonumber \\
tau_s & =& 4.98 \pm 1.07  \nonumber \\
tau_f & =& 8.31 \pm 1.51   \nonumber \\
tau_0 & =& 8.38 \pm 1.5    \nonumber \\
alpha & =& .189 \pm .004   \nonumber \\
E_0   & =& .635 \pm .072     \nonumber \\
V_0   & =& .0149 \pm .006     \nonumber 
\end{eqnarray}
%here
On the one hand, Johnston et. al. is a pseudo-maximum likelihood estimator, and 
it depends very little on the priors. On the other hand, the results are very 
different from \cite{Buxton1998}, \cite{Hu2009}, \cite{Friston2000}, and 
\cite{Friston2002}. All four of the papers just mentioned all have a common root,
and all used essentially the same priors, and all use Bayesian estimation
that depends greatly on the priors. In this case, it is possible that priors
biased the results, preventing accurate measurements of the parameters. Its also
possible that the parameters don't lend themselves to learning very well. 
It would be interesting to see what
effect different priors would have on the results in \cite{Friston2002}.
